%!TEX root = forallx-ubc.tex
%\chapter{SL Trees}
%\label{ch.sl.trees}



\practiceproblemsA{ch.sl.trees}

\vspace{-2em}

If you want additional practice, you can construct trees for any of the SL arguments and entailment claims in the exercises for the previous two chapters.

\solutions
\problempart 
\label{pr.sl.treeroot}
To evaluate each of the following claims with a tree, (a) what would you put in the root of the tree?, and (b) if the tree closes, does that show that the claim is true or false?
\begin{earg}
\item $\{P, P \eif Q, Q \eif \enot P\} \vdash{}\bot$
%(a) $\{P, P \eif Q, Q \eif \enot P\}$, (b) true
\item $(P \eif Q) \eiff (Q \eif P)$ is a tautology.
%(a) $\enot((P \eif Q) \eiff (Q \eif P))$, (b) true
\item The following argument is valid:
	\begin{ekey}
		\item[] $P \eand Q$
		\item[] $\enot R \eif \enot Q$
		\item[\therefore] $P \eand R$
	\end{ekey}
%(a) $\{P \eand Q, \enot R \eif \enot Q, \enot (P \eand R)\}$, (b) true
\item There is no interpretation that satisfies $A \eor B$, $B \eif C$, and $A \eiff C$ without also satisfying C.
%$\{A \eor B, B \eif C, A \eiff C, \enot C\}, (b) true
\item $A \eiff \enot A$ is a contradiction.
%(a) $A \eiff \enot A$, (b) false
\item Every interpretation satisfying $P$, $P \eif Q$, and $\enot Q$ also satisfies $A$.
%(a) $\{P, P \eif Q, \enot Q, \enot A\}$
\item There is at least one interpretation that satisfies $P \eif Q$, $\enot P \eor \enot Q$, and $Q \eif P$.
%(a) $\{P \eif Q, \enot P \eor \enot Q, Q \eif P\}, (b) false.
 \end{earg}

\solutions
\problempart 
\label{pr.sl.agtree}
Evaluate the argument given on p.\ \pageref{8letterargument} by constructing a tree. If it is invalid, give a model demonstrating it so.

\solutions
\problempart Evaluate each claim from Part \ref{pr.sl.treeroot} by constructing a tree. If applicable, give the interpretation that demonstrates the claim true or false.
\label{tree.examples}

\problempart Recall the discussion of the Sheffer stroke in Chapter \ref{ch.TruthTables}, page \pageref{pr.altConnectives}.  That connective, if added to SL, would have this characteristic truth table:

\begin{center}
\begin{tabular}{c|c|c}
\metaA{} & \metaB{} & \metaA{}$|$\metaB{}\\
\hline
1 & 1 & 0\\
1 & 0 & 1\\
0 & 1 & 1\\
0 & 0 & 1
\end{tabular}
\end{center}

What would be appropriate resolution rules for the Sheffer stroke, or the negated Sheffer stroke?
