% !TeX root = ./slides-05.tex

\setcounter{section}{4}

\section{Introduction to Quantifier Logic}

\subsection{The goals of QL}

\begin{frame}
  \frametitle{Limits of symbolization in SL}

\begin{itemize}[<+->]
  \item Consider the argument:
  \begin{earg}
    \item[] Greta is a hero.
    \item[\therefore] There is a hero.
  \end{earg}
  \item It's clearly valid: in any case in which Greta is a hero,
  someone (or something, at least) is a hero, so there must be a hero.
  \item But its symbolization in SL is invalid in SL:
  \begin{earg}
    \item[] $G$
    \item[\therefore] $H$
  \end{earg}
\end{itemize}
\end{frame}

\begin{frame}
  \frametitle{The problem}

  \begin{itemize}[<+->]
    \item Symbolization in SL allows us to break down sentences
    containing ``and,'' ``or,'' ``if--then'' and determine validity in
    virtue of these \emph{connectives}.
    \item Anything that can't be further broken down must be
    symbolized by a sentence letters.
    \item That includes basic sentences like ``Greta is a hero,'' but also:
    \begin{itemize}
      \item Everyone is a hero.
      \item No one is a hero.
      \item All heroes wear capes.
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{The goals of QL}
  \begin{itemize}[<+->]
    \item Finer-grained symbolization
    \item Combines with SL
    \item Allows for precise semantics (like truth tables for SL)
    \item Works with proof rules (add new rules)
    \item Be simple \& expressive (few new symbols!)
    \item New language: \emph{quantifier logic QL}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{The goals of QL}

  \begin{itemize}[<+->]
    \item Consider the valid argument:
      \begin{earg}
        \item[] Greta is a hero.
        \item[] Greta does not wear a cape.
        \item[\therefore] Not all heroes wear capes.
      \end{earg}
    \item We'll need way to connect the occurrences of the name
    ``Greta'' in the premises.
    \item We'll need to connect ``hero'' in the premise and conclusion.
    \item We want to retain use of \enot{} for ``not''
    \item We want the symbolization to have a proof.
  \end{itemize}
\end{frame}

\subsection{Beginning symbolization in QL}

\begin{frame}
  \frametitle{First steps: names}

  \begin{itemize}[<+->]
    \item Purpose of a \emph{proper name}: to pick out a single, specific thing.
    \item (Contrast with common nouns like ``hero'' or ``rock'' which pick out collections of things)
    \item For simplicity, we'll only consider names that pick out an
    actual object (in any case we're considering)
    \item Later on, we'll be able to deal with other expressions that play a similar role to names, e.g., ``the president of the USA''
    \item In QL, names will be lowercase letters $a$--$r$
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{First steps: predicates}

  \begin{itemize}[<+->]
    \item Remove a name from a sentence. What's left over is a \emph{predicate}, e.g.,\\
      \begin{earg}
        \item[] Greta is a hero
        \item[$\Rightarrow$] \gap{x} is a hero.
      \end{earg}
      \begin{earg}
        \item[] Greta admires Autumn
        \item[$\Rightarrow$] \gap{x} admires \gap{y}.
      \end{earg}
    \item In QL, predicates are symbolized using uppercase letters
    $A$--$Z$ plus a number of argument slots (marked with variables),
    e.g., $H\qv{x}$ or $A\qr{x}{y}$.
    \item Argument slots correspond to blanks.
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Symbolization keys}

  \begin{itemize}[<+->]
    \item \emph{Names}: lowercase letters for proper names of English
    \item \emph{Predicates}: uppercase letters with variables marking
    blanks.
    \begin{ekey}
    \item[a] Autumn
    \item[g] Greta
    \item[H\qv{x}] \gap{x} is a hero
    \item[V\qv{x}] \gap{x} is a villain 
    \item[I\qv{x}] \gap{x} inspires
    \item[C\qv{x}] \gap{x} wears a cape
    \item[W\qr{x}{y}] \gap{x} watched \gap{y}
    \item[A\qr{x}{y}] \gap{x} admires \gap{y}
    \item[Y\qr{x}{y}] \gap{x} is younger than \gap{y} 
    \end{ekey}
    \item \emph{Domain}: what things we're talking about\\
    e.g., people alive in \year{}
  \end{itemize}
\end{frame}

\begin{frame}
\frametitle{Symbolization}

\begin{itemize}[<+->]
\item Basic sentences: predicates with names replacing variables.
\begin{itemize}[<+->]
  \item Greta is a hero. \alert{$H\qv{g}$}
  \item Greta admires Autum. \alert{$A\qr{g}{a}$}
\end{itemize}
\item Combinations using connectives:
\begin{itemize}[<+->]
  \item Greta and Autumn are heroes. \alert{$H\qv{g} \eand H\qv{a}$}
\end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}
\frametitle{Symbolization}

\begin{itemize}[<+->]
\item Replacing pronouns by antecedents:
\begin{itemize}[<+->]
  \item If Autumn is a hero, Greta admires \textbf{her}. \alert{$H\qv{a} \eif A\qr{g}{a}$}
  \item Greta doesn't admire \textbf{herself}. \alert{$\enot A\qr{g}{g}$}
  \item Greta and Autumn watched \textbf{each other}. \alert{$W\qr{g}{a} \eand W\qr{a}{g}$}
\end{itemize}
\item Modification:
\begin{itemize}[<+->]
  \item Autumn is an \textbf{inspiring hero}.\\
  Autumn inspires and is a hero. \alert{$I\qv{a} \eand H\qv{a}$}
  \item Greta is a \textbf{hero who doesn't wear a cape}.\\
   Greta is a hero and it's not the case that Greta wears a cape.
   \alert{$H\qv{g} \eand \enot C\qv{g}$}
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Careful with modification}

\begin{itemize}[<+->]
\item Greta is an international hero.
\begin{itemize}
  \item Can't be paraphrased as\\
  ``Greta is international and a hero.''
  \item So ``\gap{} is an international hero'' needs its own
  predicate.
\end{itemize}
\item The Piltdown Man is a fake fossil.
\begin{itemize}[<+->]
  \item Can't be paraphrased as\\
  ``The Piltdown Man is fake and a fossil.''
  \item ``Fake'' and other privative adjectives (``pretend,''
  ``fictitious'') imply opposite!
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Examples}

\begin{itemize}[<+->]
  \item Autumn and Greta are inspiring heroes.\\
  \item[] \alert{$(I\qv{a} \eand H\qv{a}) \eand (I\qv{g} \eand H\qv{g})$}
  \item Greta admires Autumn but not herself.\\
  \item[] \alert{$A\qr{g}{a} \eand \enot A\qr{g}{g}$}
  \item Greta inspires only if Autumn does.
  \item[] \alert{$I\qv{g} \eif I\qv{a}$}
  \item Greta and Autumn watched each other.
  \item[] \alert{$W\qr{g}{a} \eand W\qr{a}{g}$} 
  \item Greta is older than Autumn.
  \item[] \alert{$Y\qr{a}{g}$}  
  \item One of Greta and Autumn watched the other.
  \item[] At least one: \hfill Exactly one:\\
  \alert{$W\qr{g}{a} \eor W\qr{a}{g}$} \hfill \alert{$(W\qr{g}{a} \eor W\qr{a}{g}) \eand \enot(W\qr{g}{a} \eand W\qr{a}{g})$}
\end{itemize}
\end{frame}

\subsection{The existential quantifier}

\begin{frame}
\frametitle{Existential quantifier}

\begin{itemize}[<+->]
  \item In English: ``something,'' ``someone,'' ``there is \dots''
  \item For instance:
  \begin{itemize}[<+->]
    \item \emph{Someone} wears a cape.
    \item \emph{There is} a hero.
    \item \emph{Something} inspires.
  \end{itemize}
  \item Note: can (often) go where names and pronouns also go.
  \item But works differently from names (``something'' doesn't pick
  out a unique, specific object).
\end{itemize}
\end{frame}

\begin{frame}
  \frametitle{How to symbolize ``something''}

  \begin{itemize}[<+->]
    \item Idea: introduce a special term $\mathit{sg}$?
    \item Problem: now we can't distinguish between
    \begin{itemize}[<+->]
      \item Someone is a hero and wears a cape.
      \item Someone is a hero and someone wears a cape.
    \end{itemize}
    as both would be symbolized by $H(\mathit{sg}) \eand
    C(\mathit{sg})$.
    \item Better idea: symbolize (complex) \emph{properties} and introduce
    mechanism for expressing that properties are \emph{instantiated}.
  \end{itemize}
\end{frame}

\begin{frame}
\frametitle{Expressing properties}

  \begin{itemize}[<+->]
    \item One-place predicates \emph{express} properties, e.g.,
    \begin{itemize}[<+->]
      \item $H\qv{x}$ expresses property ``being a hero.''
      \item $I\qv{x}$ expresses ``is inspiring.''
    \end{itemize}
    \item Combinations of predicates (with connectives, names) can
    express \emph{derived} properties, e.g.,
    \begin{itemize}[<+->]
      \item $A\qr{x}{g}$ expresses ``admires Greta''
      \item $W\qr{a}{x}$ expresses ``is watched by Autumn''
      \item $H\qv{x} \eand C\qv{x}$ expresses ``is a hero who wears a cape''
    \end{itemize}
    \item Note: all contain a \emph{single} variable $x$
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{The existential quantifier $\exists$}

  \begin{itemize}[<+->]
    \item Symbol for ``there is'': \alert{$\exists$}
    \item Combine $\exists$ with expression for a property (e.g., $(H\qv{x} \eand
    C\qv{x})$) to say ``something (someone) has that property''
    \item Put the variable that serves as a marker for the gap also
    after $\exists$. E.g.,
    \[ \alert{\qt{\exists}{x}\,(H\qv{x} \eand C\qv{x})} \]
    says ``Someone is a hero and wears a cape''
  \end{itemize}
\end{frame}

\begin{frame}
\frametitle{Quantifiers and variables}

  \begin{itemize}[<+->]
    \item Compare:
    \begin{align*}
    \qt{\exists}{x}\,(H\qv{x} & \eand C\qv{x}) \text{ to }\\
    \qt{\exists}{x} \, H\qv{x} &\eand \qt{\exists}{x}\, C\qv{x}
    \end{align*}
    \item In first case, the same person must be a hero and wear a cape.
    \item In second case, one person can be the hero, and another wears a cape.
    \item Multiple $\qt{\exists}{x}$ are independent, even if they use
    the same~$x$. No difference in meaning between
    \begin{align*}
     & \qt{\exists}{x} \, H\qv{x} \eand \qt{\exists}{x}\, C\qv{x} \text{ and}\\
     & \qt{\exists}{x} \, H\qv{x} \eand \qt{\exists}{y}\, C\qv{y}.
    \end{align*}
  \end{itemize}
\end{frame}

\begin{frame}
\frametitle{The domain and quantifiers}

\begin{itemize}[<+->]
  \item Symbolization key gives a domain of objects being talked about.
  \item Quantifier \emph{ranges over} this domain.
  \item That means: $\qt{\exists}{x}\, \dots x \dots$ is true iff some object
  \emph{in the domain} has the property expressed by $\dots x \dots$.
  \item Domain makes a difference: Consider $\qt{\exists}{x}\, W\qr{x}{g}$.
  \begin{itemize}
    \item True if someone watched Greta (say, Autumn did).
    \item Now take the domain to include only Greta.
    \item Relative to that domain, $\qt{\exists}{x}\, W\qr{x}{g}$ is true iff
    Greta watched herself.
  \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Quantifier restriction in English}

\begin{itemize}[<+->]
  \item ``something'' and ``someone'' work grammatically like singular
  terms (go where names can also go).
  \item ``some'' (on its own) does not: it is a \emph{determiner} and
  needs a \emph{complement}, e.g.,
  \begin{itemize}
    \item a common noun (``some hero''), or
    \item a noun phrase (``some admirer of Greta'').
  \end{itemize}
  \item ``some'' + complement works grammatically like ``someone'',
  e.g., ``\emph{Some hero} wears a cape''
  \item General form: ``Some $F$ is $G$.''
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Quantifier restriction in QL}

\begin{itemize}[<+->]
  \item ``Some $F$ is $G$'' \emph{restricts} the ``something''
  quantifier to $F$s.
  \item We could (and linguists often do) mark restrictions in the
  quantifier, e.g., $(\qt{\exists}{x}\colon F\qv{x}) G\qv{x}$
  \item We won't because we can do without.
  \item ``Some $F$ is $G$'' is true iff there is something which
  is both $F$ and also $G$, so:
  \item ``Some $F$ is $G$'' can be symbolized as
  \[ \qt{\exists}{x}(F\qv{x} \eand G\qv{x}) \]
  \item We'll also symbolize the plural form this way\\ (``Some $F$s are
  $G$s'').
  \item And more generally (most) sentences of the form:\\
  ``$G$(some $F$)'' or ``$G$(something that $F$s)''.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Examples}

\begin{itemize}[<+->]
  \item \alert{Some hero} wears a cape.\\
  \alert{Some heroes} wear capes.\\
  \item[] \alert{$\qt{\exists}{x}(H\qv{x} \eand C\qv{x})$}
  \item \alert{Someone who wears a cape} watched Greta.\\
  \item[] \alert{$\qt{\exists}{x}(C\qv{x} \eand W\qr{x}{g})$}
  \item Greta admires \alert{some hero who wears a cape}.
  \item[] \alert{$\qt{\exists}{x}((H\qv{x} \eand C\qv{x}) \eand A\qr{g}{x})$}
  \item Autumn watched \alert{someone who watched Greta}.
  \item[] \alert{$\qt{\exists}{x}(W\qr{x}{g} \eand W\qr{a}{x})$}
\end{itemize}
\end{frame}

\subsection{The universal quantifier}

\begin{frame}
\frametitle{Universal quantifier}

\begin{itemize}[<+->]
  \item ``\emph{Something} is $F$'' is true iff \emph{at least one} element of domain
  is~$F$.
  \item ``\emph{Everything} is $F$'' is true iff \emph{every element} of the domain
  is~$F$.
  \item In QL: $\qt{\forall}{x}\, F\qv{x}$.
  \item E.g.:
    \begin{itemize}
      \item ``Everyone wears a cape'' \alert{$\qt{\forall}{x}\,C\qv{x}$}
      \item ``Everyone watched Greta or Autumn'' \alert{$\forall
      x(W\qr{x}{g} \eor W\qr{x}{a})$}
    \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Universal determiners: all, every, any}

\begin{itemize}[<+->]
  \item Determiners with universal meaning: \emph{all, every, any}.
  \item Take complements (just like ``some'' does), e.g.,
      \begin{itemize}[<+->]
        \item \alert{Every hero} inspires.
        \item \alert{All heroes} inspire.
        \item \alert{Any hero} inspires.
      \end{itemize}
  \item These are true in the same cases (mean the same).
  \item ``Every $F$ is $G$'' is true iff everything \alert{which is
  $F$} is $G$.
  \item Watch out for ``any'': not always universal.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Restricted $\forall$ in QL}

  \begin{itemize}[<+->]
    \item Suppose we can symbolize $F$ and $G$.
    \item How do we symbolize ``Every $F$ is $G$''?
    \item Options:
      \begin{itemize}[<+->]
        \item \alert{$\qt{\forall}{x}(F\qv{x} \eand G\qv{x})$}
        \\\uncover<7->{If true, everything must be $F$.\\
        So can be false when ``Every $F$ is $G$'' is true.}
        \item \alert{$\qt{\forall}{x}(F\qv{x} \eor G\qv{x})$}
        \\ \uncover<8->{True if everything is $F$ (without being $G$).\\
        So can be true when ``Every $F$ is $G$'' is false.}
        \item \alert{$\qt{\forall}{x}(F\qv{x} \eif G\qv{x})$}
        \\ \uncover<9>{If $x$ is $F$, $x$ must also be $G$.\\
        (If $x$ is not $F$, doesn't matter if it's $G$ or not.)}
      \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}
\frametitle{Symbolizing ``all $F$s are $G$s''}

Symbolize the following as \emph{\[\qt{\forall}{x}(F\qv{x} \to G\qv{x})\]}

\begin{itemize}
  \item All $F$s are $G$s.
  \item Every $F$ is $G$.
  \item Any $F$ is $G$.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Examples}

\begin{itemize}[<+->]
  \item \emph{Every hero} wears a cape.\\
  \emph{All heroes} wear capes.\\
  \item[] \alert{$\qt{\forall}{x}(H\qv{x} \eif C\qv{x})$}
  \item \emph{Every hero who wears a cape} watched Greta.\\
  \item[] \alert{$\qt{\forall}{x}((H\qv{x} \eand C\qv{x}) \eif W\qr{x}{g})$}
  \item Greta and Autumn admire \emph{anyone who wears a cape}.
  \item[] \alert{$\qt{\forall}{x}(C\qv{x} \eif (A\qr{g}{x} \eand A\qr{a}{x}))$}
  \item Autumn watched \emph{everyone who watched Greta}.
  \item[] \alert{$\qt{\forall}{x}(W\qr{x}{g} \eif W\qr{a}{x})$}
  \item All heroes and villains watched Greta. 
  \item[] \alert{$\qt{\forall}{x}((H\qv{x}
  \mathbin{\colorbox{highlightbg}{\eor}} V\qv{x}) \eif W\qr{x}{g})$} 
\end{itemize}
\end{frame}

\subsection{No, only, a, and some \& any again}

\begin{frame}
\frametitle{No $F$ is $G$}

\begin{itemize}[<+->]
  \item ``\emph{No $F$s are $G$s}'' can be paraphrased as 
\begin{itemize}[<+->]
  \item ``\emph{Every} $F$ is \emph{not}-$G$,'' or as
  \item ``\emph{Not: some} $F$ is $G$.''
  \end{itemize}
  \item So symbolize it using:
\begin{itemize}
  \item \emph{$\qt{\forall}{x}(F\qv{x} \to \enot G\qv{x})$} or
  \item \emph{$\enot\qt{\exists}{x}(F\qv{x} \eand G\qv{x})$}
\end{itemize}
\end{itemize}

\end{frame}

\begin{frame}
\frametitle{Examples}

\begin{itemize}[<+->]
  \item \emph{No hero} wears a cape.\\
  \emph{No heroes} wear capes.\\
  \item[] \alert{$\qt{\forall}{x}(H\qv{x} \eif \enot C\qv{x})$}
  \item \emph{No hero who wears a cape} watched Greta.\\
  \item[] \alert{$\qt{\forall}{x}((H\qv{x} \eand C\qv{x}) \eif \enot W\qr{x}{g})$}
  \item Greta admires \emph{no one who wears a cape}.
  \item[] \alert{$\enot\qt{\exists}{x}(C\qv{x} \eand A\qr{g}{x})$}
  \item Autumn watched \emph{no one who watched Greta}.
  \item[] \alert{$\enot\qt{\exists}{x}(W\qr{x}{g} \eand W\qr{a}{x})$}
\end{itemize}
\end{frame}


\begin{frame}
\frametitle{Only $F$s are $G$}

\begin{itemize}[<+->]
\item When is ``Only $F$s are $G$s'' false?
\item When there is a \emph{non}-$F$ that's $G$.
\item So symbolize it as
\[
  \alert{\enot\qt{\exists}{x}(\enot F\qv{x} \eand G\qv{x})}
  \]
\item Or, paraphrase it as: ``Any $x$ is $G$ \emph{only if} it is $F$''
\item So another symbolization is:
\[
  \alert{\qt{\forall}{x}(G\qv{x} \to F\qv{x})}
\]
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Examples}

\begin{itemize}[<+->]
  \item Only heroes wear capes.\\
  \item[] \alert{$\qt{\forall}{x}(C\qv{x} \eif H\qv{x})$}
  \item Only heroes who wear capes watched Greta.\\
  \item[] \alert{$\qt{\forall}{x}(W\qr{x}{g} \eif (H\qv{x} \eand C\qv{x}))$}
  \item Greta admires only people who wear capes.
  \item[] \alert{$\enot\qt{\exists}{x}(\enot  C\qv{x} \eand A\qr{g}{x})$}
  \item Autumn watched only heroes and villains.
  \item[] \alert{$\enot\qt{\exists}{x}(\enot (H\qv{x} \eor V\qv{x}) \eand W\qr{a}{x})$}
  \item[] \alert{$\qt{\forall}{x}(W\qr{a}{x} \eif (H\qv{x} \eor V\qv{x}))$}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{The indefinite article}

\begin{itemize}[<+->]
  \item We use ``is a'' to indicate predication, e.g., ``Greta is a hero.''
  \item Often ``a'' is used to claim existence, e.g.,
  \begin{itemize}[<+->]
    \item[] Greta admires a hero.
    \item[] \alert{$\qt{\exists}{x} (H\qv{x} \eand A\qr{g}{x})$}
  \end{itemize}
  \item But a \emph{generic} indefinite is closer to a
  universal quantifier:
  \begin{itemize}[<+->]
    \item[] A hero is someone who inspires.
    \item[] \alert{$\qt{\forall}{x}(H\qv{x} \eif I\qv{x})$}
  \end{itemize}
  \item Be careful if the indefinite article is in the antecedent of a conditional:
  \begin{itemize}[<+->]
    \item[] If \emph{a hero} wears a cape, \emph{they} inspire.
    \item[] That means: all heroes who wear capes inspire.
    \item[] \alert{$\qt{\forall}{x}((H\qv{x} \eand C\qv{x}) \eif I\qv{x})$}
  \end{itemize}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Universal ``some'', existential ``any''}

\begin{itemize}[<+->]
  \item ``Someone,'' ``something'' can require a \emph{universal} quantifier:\\
 if it's in the antecedent of a conditional, with a pronoun in the consequent referring back to it, e.g.,
  \begin{itemize}[<+->]
    \item[] If \emph{someone} is a hero, Autum admires \emph{them}.
    \item[] Roughly: Autumn admires all heroes.
    \item[] \alert{$\qt{\forall}{x}(H\qv{x} \eif A\qr{a}{x})$}
  \end{itemize}
  \item ``Any'' in antecedents but \emph{without} pronouns referring back to
  them are \emph{existential}:
  \begin{itemize}[<+->]
    \item[] If \emph{anyone} is a hero, Greta is.
    \item[] Roughly: if there are heroes (at all), Greta is a hero.
    \item[] \alert{$\qt{\exists}{x}\, H\qv{x} \eif H\qv{g}$}
  \end{itemize}
\end{itemize}
\end{frame}

\subsection{Mixed domains}

\begin{frame}
\frametitle{Mixed domains}

\begin{itemize}[<+->]
  \item Sometimes you want to talk about more than one kind of thing.
  \item The domain can include any mix of things (e.g., people,
  animals, items of clothing)
  \item Proper symbolization then needs predicates for these kinds, e.g.:
    \begin{ekey}
    \item[$Domain$] people alive in \year{} and items of clothing
    \item[P\qv{x}] \gap{x} is a person
    \item[L\qv{x}] \gap{x} is an item of clothing.
    \item[E\qv{x}] \gap{x} is a cape
    \item[R\qr{x}{y}] \gap{x} wears \gap{y}
    \end{ekey}
\end{itemize}
\end{frame}


\begin{frame}
\frametitle{Quantification in mixed domains}

\begin{itemize}[<+->]
\item Not everyone is wearing a cape.
\begin{itemize}
  \item In domain of people only:
  \item[] \alert{$\enot \qt{\forall}{x}\, C\qv{x}$}
  \item In mixed domain:
  \item[] \alert{$\enot \qt{\forall}{x}\, (P\qv{x} \eif C\qv{x})$}
\end{itemize}
\item Some people inspire.
\begin{itemize}[<+->]
  \item In domain of people only:
  \item[] \alert{$\qt{\exists}{x}\,I\qv{x}$}
  \item In mixed domain:
  \item[] \alert{$\qt{\exists}{x}(P\qv{x} \eand I\qv{x})$}
\end{itemize}
\item Greta wears something.
\item[] \alert{$\qt{\exists}{x}(L\qv{x} \eand R\qr{g}{x})$}
\end{itemize}
\end{frame}

