% !TeX root = ./slides-07.tex

\setcounter{section}{6}

\section{Proofs in QL}

\subsection{Rules for $\forall$}

\begin{frame}
\frametitle{Rules for formal proofs}

\begin{itemize}[<+->]
\item Need rules for $\forall$ and $\exists$ for formal proofs
\item Formal proofs now more important, because no alternative
  (truth-table method)
\item Intro and Elim rules should be
\begin{itemize}[<+->]
\item simple
\item elegant (not involve other connectives or quantifiers)
\item yield only valid arguments
\end{itemize}
\end{itemize}

\end{frame}

\begin{frame}
\frametitle{Candidates for rules}

\begin{itemize}[<+->]
\item Only simple sentence close to $\qt{\forall}{x}\, \metav A\qv{\metav{x}}$ is
$\metav A\qv{\metav{c}}$
\item Gives simple, elegant $\forall$E rule:

\begin{fitchproof}
\have[k]{a}{\qt{\forall}{\metav{x}}\, \metav A\qv{\metav{x}}}
\have[ ]{b}{\metav A\qv{\metav{c}}}\Ae{a}
\end{fitchproof}
\item This is a good rule: $\qt{\forall}{\metav{x}}\,\metav A\qv{\metav{x}}
\models \metav A\qv{\metav{c}}$.
\end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Candidates for rules}

\begin{itemize}[<+->]
\item Problem: corresponding ``intro rule'' isn't valid:
\begin{fitchproof}
\have[k]{a}{\metav A\qv{\metav{c}}}
\have[ ]{b}{\qt{\forall}{\metav{x}}\,\metav A\qv{\metav{x}}}\by{\emph{doesn't follow from}}{a}
\end{fitchproof}
\item Diagnosis: the $\metav c$ in $\metav A\qv{\metav{c}}$ is a name for
a \emph{specific object}.
\item We need a name for an \emph{arbitrary,
unspecified object}.
\item If $\metav A\qv{\metav{c}}$ is true for whatever $\metav c$
\emph{could} name, then $\metav A\qv{\metav{x}}$ is satisfied by \emph{every} object.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Names for arbitrary objects}

\begin{itemize}[<+->]
\item When we give proofs of general claims, we often do use names for
arbitrary objects (well, mathematicians do at least).

\begin{earg}
\item[] All heroes admire Greta.
\item[] Only people who wear capes admire Greta.
\item[\therefore] All heroes wear capes.
\end{earg}
\pause
Proof: Let Carl be any hero.  Since all heroes admire Greta, Carl
admires Greta. Since only people who wear capes admire Greta, Carl is
wears a cape. But ``Carl'' stands for \emph{any} hero. So all heroes
wear capes.
\end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Universal generalization}

\begin{fitchproof}
  \have[k]{a}{\metav{A}\qv{\metav{c}}}
  \have[ ]{b}{\qt{\forall}{\metav{x}}\,\metav{A}\qv{\metav{x}}}\Ai{a}
\end{fitchproof}
  \begin{itemize}[<+->]
  \item $\metav{c}$ is special: $\metav{c}$ must not appear in any
  premise or assumption of a subproof not already ended
  \item $\metav{A}\qv{\metav{x}}$ is obtained from $\metav{A}\qv{\metav{c}}$ by replacing \emph{all} occurrences of $\metav{c}$ by $\metav{x}$.
  \item In other words, $\metav{c}$ must also not occur in $\forall
  \metav{x}\,\metav{A}\qv{\metav{x}}$.
 \end{itemize}
\end{frame}

\begin{frame}
\frametitle{General conditional proof}

Proving ``All $A$s are $B$s''

\begin{fitchproof}
  \open
  \hypo[k]{a}{A\qv{c}}
  \have[l]{b}{B\qv{c}}
  \close
  \have{c}{A\qv{c} \eif B\qv{c}}\ci{a-b}
  \have{d}{\qt{\forall}{x}(A\qv{x} \eif B\qv{x})}\Ai{c}
\end{fitchproof}
\end{frame}

\begin{frame}
\frametitle{Example}

\begin{earg}
  \item[] All heroes admire Greta.
  \item[] Only people who wear capes admire Greta.
  \item[\therefore] All heroes wear capes.
\end{earg}

\bigskip

\begin{earg}
  \item[] $\qt{\forall}{x}(H\qv{x} \eif A\qr{x}{g})$
  \item[] $\qt{\forall}{x}(A\qr{x}{g} \eif C\qv{x})$
  \item[\therefore] $\qt{\forall}{x}(H\qv{x} \eif C\qv{x})$
\end{earg}

Let's do it on \href{https://carnap.io/shared/rzach@ucalgary.ca/Practice\%20Problems\%20VII.md}{carnap.io}
\end{frame}

\begin{frame}
\frametitle{Example}
\small
\begin{fitchproof}
  \hypo{a}{\qt{\forall}{x}(H\qv{x} \eif A\qr{x}{g})}
  \hypo{b}{\qt{\forall}{x}(A\qr{x}{g} \eif C\qv{x})}
  \open
  \hypo{c}{H\qv{c}}
  \have{d}{H\qv{c} \eif A\qr{c}{g}}\Ae{a}
  \have{e}{A\qr{c}{g}}\ce{d,c}
  \have{f}{A\qr{c}{g} \eif C\qv{c}}\Ae{b}
  \have{g}{C\qv{c}}\ce{f,e}
  \close
  \have{h}{H\qv{c} \eif C\qv{c}}\ci{c-g}
  \have{i}{\qt{\forall}{x}(H\qv{x} \eif C\qv{x})}\Ai{h}
\end{fitchproof}
\end{frame}

\begin{frame}
  \frametitle{Example}
  \small
  \begin{fitchproof}
    \hypo{1}{\qt{\forall}{x}\,A\qv{x} \eor \qt{\forall}{x}\, B\qv{x}}
    \open
    \hypo{2}{\qt{\forall}{x}\,A\qv{x}}
    \have{3}{A\qv{c}}\Ae{2}
    \have{4}{A\qv{c} \eor B\qv{c}}\oi{3}
\close
\open
\hypo{5}{\qt{\forall}{x}\,B\qv{x}}
\have{6}{B\qv{c}}\Ae{5}
\have{7}{A\qv{c} \eor B\qv{c}}\oi{6}
\close
  \have{8}{A\qv{c} \eor B\qv{c}}\oe{1,2-4,5-7}
  \have{9}{\qt{\forall}{x}(A\qv{x} \eor B\qv{x})}\Ai{8}
  \end{fitchproof}
  \end{frame}

\subsection{Rules for $\exists$}

\begin{frame}
\frametitle{Rules for $\exists$}

\begin{itemize}[<+->]
\item If we know of a specific object that it satisfies $\metav A\qv{x}$, we know that at least one object satisfies $\metav A\qv{x}$.
\item So this rule is valid:
\begin{fitchproof}
  \have[k]{a}{\metav A\qv{\metav{c}}}
  \have[ ]{b}{\qt{\exists}{\metav{x}}\, \metav A\qv{\metav{x}}}\Ei{a}
\end{fitchproof}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Arbitrary objects again}

\begin{itemize}
  \item Problem: corresponding ``elim rule'' isn't valid:
  \begin{fitchproof}
    \have[k]{a}{\qt{\exists}{\metav{x}}\, \metav A\qv{\metav{x}}}
    \have[ ]{b}{\metav A\qv{\metav{c}}}\by{\emph{doesn't follow from}}{a}
  \end{fitchproof}
  \item If we know that $\qt{\exists}{\metav{x}}\, \metav A\qv{\metav{x}}$ is
  true, we know that \emph{some} objects satisfy $\metav A\qv{\metav{x}}$,
  but not which ones.
\item To use this information, we have to introduce a temporary name
that stands for any one of the objects that satisfy $\metav A(\metav
x)$.
\end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Reasoning from existential information}

  \begin{itemize}
    \item To use $\qt{\exists}{\metav{x}}\,\metav A\qv{\metav{x}}$, pretend the
    $\metav x$ has a name~$\metav c$, and reason from $\metav A(\metav
    c)$.
    \item This is what we'd do if we reason informally from existential information, e.g.,
  \begin{earg}
  \item[] There are heroes who wear capes.
  \item[] Anyone who wears a cape admires Greta.
  \item[\therefore] Some heroes admire Greta.
  \end{earg}

\pause
Proof: We know there are heroes who wear capes. Let Cate be an
arbitrary one of them.  So Cate wears a cape. Since anyone who wears a
cape admires Greta, Cate admires Greta. Since Cate is a hero who
admires Greta, some heroes admire Greta.
\end{itemize}

\end{frame}

\begin{frame}
\frametitle{Existential elimination}

\begin{itemize}[<+->]
  \item If
  \begin{itemize}[<+->]
    \item we know that some object satisfies $\metav A\qv{\metav{x}}$,
    \item we assume for the time being that $c$ is one of them (i.e.,
  assume $\metav A\qv{\metav{c}}$), and
    \item we can prove that $\metav B$ follows from this assumption,
\end{itemize}
\item[] then $\metav B$ follows already from $\qt{\exists}{\metav{x}}\, \metav A\qv{\metav{x}}$.
\item Rule for existential elimination:

\begin{fitchproof}
  \have[k]{a}{\qt{\exists}{\metav{x}}\, \metav A\qv{\metav{x}}}
  \open
  \hypo[m]{b}{\metav A\qv{\metav{c}}}
  \have[n]{c}{\metav B}
\close
\have[ ]{d}{\metav B}\Ee{a,b-c}
\end{fitchproof}
\item $\metav c$ is special: $\metav c$ must not appear outside subproof
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Example}

\begin{earg}
  \item[] There are heroes who wear capes.
  \item[] Anyone who wears a cape admires Greta.
  \item[\therefore] Some heroes admire Greta.
\end{earg}
\bigskip
\begin{earg}
  \item[] $\qt{\exists}{x}(H\qv{x} \eand C\qv{x})$
  \item[] $\qt{\forall}{x}(C\qv{x} \eif A\qr{x}{g}$
  \item[\therefore] $\qt{\exists}{x}(H\qv{x} \eand A\qr{x}{g})$
\end{earg}
\end{frame}

\begin{frame}
\frametitle{Example}
\small
\begin{fitchproof}
  \hypo{a}{\qt{\exists}{x}(H\qv{x} \eand C\qv{x})}
  \hypo{b}{\qt{\forall}{x}(C\qv{x} \eif A\qr{x}{g}}
  \open
  \hypo{c}{H\qv{c} \eand C\qv{c}}
  \have{d}{C\qv{c}}\ae{c}
  \have{e}{C\qv{c} \eif A\qr{c}{g}}\Ae{b}
  \have{f}{A\qr{c}{g}}\ce{d,e}
  \have{g}{H\qv{c}}\ae{c}
  \have{h}{H\qv{c} \eand A\qr{c}{g}}\ai{d,g}
  \have{i}{\qt{\exists}{x}(H\qv{x} \eand A\qr{x}{g})}\Ei{h}
  \close
  \have{j}{\qt{\exists}{x}(H\qv{x} \eand A\qr{x}{g})}\Ei{a,c-i}
\end{fitchproof}
\end{frame}

