\documentclass[a4paper, 11pt]{article} % Font size (can be 10pt, 11pt or 12pt) and paper size (remove a4paper for US letter paper)
\usepackage[protrusion=true,expansion=true]{microtype} % Better typography
\usepackage{bibentry}
\usepackage[round]{natbib} %%Or change 'round' to 'square' for square backers
\setcitestyle{aysep={}}

\makeatletter
\renewcommand{\maketitle}{
\begin{flushright}
{\LARGE\@title}

\vspace{10pt}

{\@author}
\\ \@date
\end{flushright}

\vspace{-20pt}

}
\makeatother

%----------------------------------------------------------------------------------------
%	TITLE
%----------------------------------------------------------------------------------------

\title{\textbf{Bibliography}} % Subtitle

\author{\textsc{Logic I}\\ \em Benjamin Brast-McKie} % Institution

\date{\today} % Date

%----------------------------------------------------------------------------------------

\begin{document}

\maketitle % Print the title section

\thispagestyle{empty}

%----------------------------------------------------------------------------------------

\section*{Philosophical Logic}%
  \label{sec:PhilosophicalLogic}

  \begin{enumerate}
    \item \cite{Boolos1999} -- essays in philosophical logic and the philosophy of mathematics (some more advanced than others).
  \end{enumerate}




\section*{Modal Logic}%
  \label{sec:ModalLogic}

  \begin{enumerate}
    \item \cite{Sider2010} -- logical toolkit for philosophy (accessible).
    \item \cite{Hughes1996} -- classic text in modal logic (intermediate).
    \item \cite{Blackburn2002} -- modern text in modal logic (advanced).
    % \item \cite{Williamson2013} -- philosophy of modal logic (intermediate).
  \end{enumerate}



\section*{Mathematical Logic}%
  \label{sec:MathematicalLogic}

  \begin{enumerate}
    \item \cite{Enderton2001} -- classic introduction to metalogic (intermediate).
    \item \cite{Boolos2007} -- computability and incompleteness (intermediate).
    \item \cite{Mendelson2015} -- extensive textbook in metalogic (intermediate).
  \end{enumerate}




\section*{Type Theory and Lambda Calculus}%
  \label{sec:TypeTheory}

  \begin{enumerate}
    \item \cite{hindley2008a} -- austere treatment of type theory (advanced).
    \item \cite{ hindley2008} -- textbook in lambda calculus (advanced).
  \end{enumerate}




\section*{Higher-Order Logic}%
  \label{sec:HigherOrderLogic}
  
  \begin{enumerate}
    \item \cite{Bacon2023} -- philosophical take on higher-order logic (intermediate).
  \end{enumerate}




\section*{Plural Logic}%
  \label{sec:PluralLogic}
  
  \begin{enumerate}
    \item \cite{Oliver2013} -- introduction to plural logic (intermediate).
  \end{enumerate}



\section*{Nonclassical Logic}%
  \label{sec:Nonclassical Logic}

  \begin{enumerate}
    \item \cite{Priest2008} -- textbook covering nonclassical logics (intermediate).
  \end{enumerate}



\bibliographystyle{Phil_Review} %%bib style found in bst folder, in bibtex folder, in texmf folder.
\setlength{\bibsep}{5pt} %%Changes spacing between bib entries
\bibliography{Zotero} %%bib database found in bib folder, in bibtex folder

\end{document}


